---
title: "Splines"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read the Data
```{r}
data <- as.matrix(read.table("P04.csv",sep=",",stringsAsFactors = FALSE, header=FALSE, col.names
=c("year", "energy")))

head(data)
```


### Create the X,Y vectors from data
```{r}
Y = data[1:69,2]
X = seq(1,length(Y))
```


### Plot the data
```{r}
plot(X,Y)
```


# A) Cubic spline (d.o.f = 4+k)


### LOOCV to find best K (that minimizes MSE)
```{r}
h1 = rep(1,length(X))
h2 = X
h3 = X^2
h4 = X^3


mse = rep(0, 15)
for(k in 6:15)
{
  H = cbind(h1, h2, h3, h4)
  k_all = seq(1,length(X),length.out = (k+2))[2:(k+1)]
  
  for(j in 1:k)
  {
    h_j = (X-k_all[j])^3
    h_j[h_j <= 0] = 0
    H = cbind(H, h_j)
  }
  
  ##LOOCV
  error = rep(0, length(X)-1)
  
  for(i in 1:length(X))
  {
    X1=X[-i]
    Y1=Y[-i]
    H1=H[-i,] 
    B1=solve(t(H1)%*%H1)%*%t(H1)%*%Y1
    error[i]= Y[i]-(H[i,]%*%B1)
  }
  
  
  mse[k]=sum(error^2)/length(X)
  

}
k=seq(6,15)
MSE=mse[6:15]
plot(k,MSE ,type='l', col = "red")
```

### Report min MSE and best k
```{r}
print(paste("Min MSE = ", min(MSE) ))
print(paste("Best K = ", k[which.min(MSE)] ))

```

### Apply the cubic Spline with best K (k=7)
```{r}
k=7

h1 = rep(1,length(X))
h2 = X
h3 = X^2
h4 = X^3
H = cbind(h1, h2, h3, h4)

k_all = seq(1,length(X),length.out = (k+2))[2:(k+1)]

for(j in 1:k)
{
  h_j = (X-k_all[j])^3
  h_j[h_j <= 0] = 0
  H = cbind(H, h_j)
}

yhat=H %*% solve(t(H)%*%H)%*%t(H)%*%Y

#sigma2 = (1/(69-k))*t(Y-yhat)%*%(Y-yhat)
#yn = yhat-3*sqrt(diag(as.numeric(sigma2)*H%*%solve(t(H)%*%H)%*%t(H)))
#yp = yhat+3*sqrt(diag(as.numeric(sigma2)*H%*%solve(t(H)%*%H)%*%t(H)))
plot(X,Y,col = "red")
#lines(X,yn,col = "blue")
#lines(X,yp,col = "blue")
lines(X,yhat,col = "black")

```


# B) B-spline

```{r}
library(splines)
```


### LOOCV to find best K (that minimizes MSE)
```{r}
mse = rep(0, 15)
for(k in 6:15)
{
  nbasis=k+3-2   #nbreaks + order - 2 ; order=3 for quadrtic
  k_all = seq(1,length(X),length.out = k)
  
  B = bs(X, knots = k_all, degree = 2,intercept = FALSE)[,1:nbasis]

  ##LOOCV
  error = rep(0, length(X)-1)
  
  for(i in 1:length(X))
  {
    X1=X[-i]
    Y1=Y[-i]
    B_new=B[-i,]
    BS = solve(t(B_new)%*%B_new)%*%t(B_new)%*%Y1
    error[i]= Y[i]-(B[i,]%*%BS)
  }
  
  mse[k]=sum(error^2)/length(X)
  
}

k=seq(6,15)
MSE=mse[6:15]
plot(k,MSE ,type='l', col = "red")


```

### Report min MSE and best k
```{r}
print(paste("Min MSE = ", min(MSE) ))
print(paste("Best K = ", k[which.min(MSE)] ))

```

### Apply the cubic Spline with best K (k=11)
```{r}
k=11

# nbreaks + order - 2, order=3 for quadrtic
nbasis=k+3-2
k_all = seq(1,length(X),length.out = k)

B = bs(X, knots = k_all, degree = 2,intercept = FALSE)[,1:nbasis]
# Least square estimation
yhat = B%*%solve(t(B)%*%B)%*%t(B)%*%Y

plot(X,Y,col = "red")
lines(X,yhat,col = "black")
  
```



# C) Smoothing Splines
```{r}
# lambda = (df = n)

lambdas=seq(2e-6,1e-4, length.out = 138)
mse=rep(0, 100)
index=1

for (l in lambdas) {
  
  error = rep(0, length(X)-1)
  for(i in 1:length(X))
  {
    X1=X[-i]
    Y1=Y[-i]
    k=length(X1)
    sm = smooth.spline(X1,Y1,nknots=k, lambda=l)
    error[i]= Y[i]-(predict(sm,X[i])$y)
  }
 
  mse[index]=sum(error^2)/length(X)
  index=index+1
  
}
plot(lambdas,mse, type='l')
best_lambda = lambdas[which.min(mse)]
```


### Report min MSE and best lambda
```{r}
print(paste("Min MSE = ", min(mse) ))
print(paste("Best K = ", lambdas[which.min(mse)] ))
```


### Apply smoothing spline for best value of lambda (min MSE) lambda = 1.27e-05
```{r}
sm = smooth.spline(Y, nknots=k, lambda=best_lambda)
plot(X,Y,col = "red",lwd=3)
lines(X,sm$y,col = "black",lwd=3) 
```

# D) Kernel smoother

```{r}
# kernel function
kerf = function(z){exp(-z*z/2)/sqrt(2*pi)}

# leave-one-out CV
h1=seq(1,4,0.1)
er = rep(0, length(Y))
mse = rep(0, length(h1))
for(j in 1:length(h1))
{
  h=h1[j]
  for(i in 1:length(X))
  {
    X1=X[-i];
    Y1=Y[-i];
    z=kerf((X[i]-X1)/h)
    yke=sum(z*Y1)/sum(z)
    er[i]=Y[i]-yke
  }
  mse[j]=sum(er^2)
}
plot(h1,mse,type = "l")
h = h1[which.min(mse)]

```


### Report min MSE and best lambda
```{r}
print(paste("Min MSE = ", min(mse) ))
print(paste("Best K = ", h1[which.min(mse)] ))
```

### Apply smoother for best lambda value lambda = 1
```{r}
N=138
xall = seq(min(X),max(X),length.out = N)
f = rep(0,N);
for(k in 1:N)
{
  z=kerf((xall[k]-X)/h)
  f[k]=sum(z*Y)/sum(z);
}
ytrue = sin(xall/10)+(xall/50)^2
plot(X,Y,col = "black")
lines(xall,ytrue,col = "red")
lines(xall, f, col = "blue")
```








